Analysis of NLA17 Dissolved Gas Indicator

### 2019-12-10 scripts/dgIndicatorAnalysis.Rmd

---
title: "Analysis of NLA17 Dissolved Gas Indicator"
authors: "J. Beaulieu, R. Martin, and M. McManus"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    toc: true
    toc_float: false
    fig_caption: yes
    depth: 2
    number_sections: true
editor_options: 
  chunk_output_type: console
---

```{r results='hide', message=FALSE, warning=FALSE}
# load libraries
library(rgdal) # spatial data
library(sf) # spatial data
library(gstat) # lagged scatterplot
library(spdplyr) # apply dplyr verbs to sp objects
library(rmarkdown)
library(tidyverse)
```

@@

The dissolved gas data were currated the 'NLA' dissolved gas project in RStudio.  The code can be found at a private github repository (https://github.com/USEPA/NLA).  After aggregating across duplicate samples, the data were written to nla17gasDataAggregated_2019-10-25.txt.  A copy of this data file is stored in documents library associated with the 'ORD NLA17 Dissolved Gas' Private Group on SharePoint (https://usepa.sharepoint.com/sites/ORD_NLA17_Dissolved_Gas).  When synced to a local computer, the documents library can be read directly with R, after identifying the local directory path.
## Getting data and filtering
```{r}
# Identify local path for each user
localPath <- Sys.getenv("USERPROFILE")

# Read data file
dg <- read.table(file = paste0(localPath, 
                               "/Environmental Protection Agency (EPA)/",
                               "ORD NLA17 Dissolved Gas - Documents/",
                               "inputData/nla17gasDataAggregated_2019-11-12.txt"), 
                 header = TRUE, sep = "\t", as.is = TRUE)
```
Concentrations of CO~2~, CH~4~, and N~2~O in ambient air over the waterbodies were measured and used in the dissolved gas concentrations.  These data are not needed in subsequent analysis and will be removed from the dataframe.
```{r}
dg <- dg %>% filter(sample.source == "DG") # filter out dissolved gas 'DG' samples
```
The dataframe also contains observations from site in Alaska.  These sites were not an official part of the NLA and will not be included in this analysis.
```{r}
dg <- dg %>% filter(!(grepl("AK", site.id))) # exlclude alaska
```

The data file contains `r dg %>% distinct(site.id, visit.no) %>% summarize(n = n()) %>% pull()` unique sampling trips defined by unique combinations of 'site.id' and 'visit.no'.  There were `r dg %>% distinct(site.id) %>% summarize(n = n()) %>% pull()` waterbodies sampled, `r dg %>% filter(visit.no == 2) %>% summarize(n = n()) %>% pull()` of which were sampled twice. The primary response variables are the concentrations of CO~2~, CH~4~, and N~2~O dissolved in the surface water of each waterbody.  Concentration is expressed as both mol/L (dissolved.co2, dissolved.ch4, dissolved.n2o) and as a saturation ratio (co2.sat.ratio, ch4.sat.ratio, n2o.sat.ratio).  Saturation ratio is defined as the ratio of the measured concentration to that expected if the waterbody were in equilibrium with the atmosphere.  A value of 1 indicates equilibrium, values < 1 indicate undersaturation and values > 1 indicate supersaturation.  Undersaturated waterbodies function as a sink for the atmospheric gas whereas supersaturated waterbodies function as a source.

@@

# Check CRS
st_crs(ecoR) # no EPSG code
ecoR <- st_transform(ecoR, 5070) # convert to CONUS Albers
st_crs(ecoR) # 5070

# head(ecoR, 9) throws Error: C stack usage  30746119 is too close to the limit
# eventuall get the head output
# Bubble plots
ggplot() +
  geom_sf(data = ecoR, color = NA, aes(fill = WSA9_NAME)) +
  geom_sf(data = dg.sf, aes(size = dissolved.co2)) +
  ggtitle("Dissolved CO2")

ggplot() +
  geom_sf(data = ecoR, color = NA, aes(fill = WSA9_NAME)) +
  geom_sf(data = dg.sf, aes(size = dissolved.ch4)) +
  ggtitle("Dissolved CH4")

ggplot() +
  geom_sf(data = ecoR, color = NA, aes(fill = WSA9_NAME)) +
  geom_sf(data = dg.sf, aes(size = dissolved.n2o)) +
  ggtitle("Dissolved N2O")
```

@@

>>>>

---
title: "Analysis of NLA17 Dissolved Gas Indicator"
authors: "J. Beaulieu, R. Martin, and M. McManus"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    toc: true
    toc_float: true
    fig_caption: yes
    depth: 2
    number_sections: true
    code_folding:  hide
editor_options: 
  chunk_output_type: console
---

```{r results='hide', message=FALSE, warning=FALSE}
# load libraries
library(sf) # spatial data
library(gstat) # lagged scatterplot, variogram
#library(spdplyr) # apply dplyr verbs to sp objects.  gstat functions require sp objects
library(tidyverse) # dplyr, ggplot

# Identify local path for each user
localPath <- Sys.getenv("USERPROFILE")

# Define helper functions
# standardized formatting for column names
toEPA <- function(X1){
  names(X1) = tolower(names(X1))
  names(X1) = gsub(pattern = c("\\(| |#|)|/|-|\\+|:|_"), replacement = ".", x = names(X1))
  X1
}
```

@@

The dissolved gas data were currated the 'NLA' dissolved gas project in RStudio.  The code can be found at a private github repository (https://github.com/USEPA/NLA).  After aggregating across duplicate samples, the data were written to nla17gasDataAggregated_2019-10-25.txt.  A copy of this data file is stored in documents library associated with the 'ORD NLA17 Dissolved Gas' Private Group on SharePoint (https://usepa.sharepoint.com/sites/ORD_NLA17_Dissolved_Gas).  When synced to a local computer, the documents library can be read directly with R, after identifying the local directory path.
## Getting data and filtering

The dissolved gas data were currated in the 'NLA' dissolved gas project in RStudio.  The code can be found at a private github repository (https://github.com/USEPA/NLA).  After aggregating across duplicate samples, the data were written to nla17gasDataAggregated_2019-10-25.txt.  A copy of this data file is stored in the documents library associated with the 'ORD NLA17 Dissolved Gas' Private Group on SharePoint (https://usepa.sharepoint.com/sites/ORD_NLA17_Dissolved_Gas).  When synced to a local computer, the documents library can be read directly with R.

```{r}
# Read data file
dg <- read.table(file = paste0(localPath, 
                               "/Environmental Protection Agency (EPA)/",
                               "ORD NLA17 Dissolved Gas - Documents/",
                           "inputData/nla17gasDataAggregated_2019-11-12.txt"), 
                 header = TRUE, sep = "\t", as.is = TRUE) %>%
  # unit conversion
  mutate(dissolved.ch4 = dissolved.ch4 * 1000000, # mol/L -> umol/L
         dissolved.co2 = dissolved.co2 * 1000000, # mol/L -> umol/L,
         dissolved.n20 = dissolved.n2o * 1000000000, # mol/L -> nmol/L
         # add source/sink column
         co2.src.snk = ifelse(co2.sat.ratio > 1, "source", "sink"),
         ch4.src.snk = ifelse(ch4.sat.ratio > 1, "source", "sink"),
         n2o.src.snk = ifelse(n2o.sat.ratio > 1, "source", "sink"))

```
Concentrations of CO~2~, CH~4~, and N~2~O in ambient air over the waterbodies were measured and used in the dissolved gas concentrations.  These data are not needed in subsequent analysis and will be removed from the dataframe.
```{r}
dg <- dg %>% filter(sample.source == "DG") # select dissolved gas 'DG' samples
```
The dataframe also contains observations from site in Alaska.  These sites were not an official part of the NLA and will not be included in this analysis.
```{r}
dg <- dg %>% filter(!(grepl("AK", site.id))) # exlclude alaska
```

The data file contains `r dg %>% distinct(site.id, visit.no) %>% summarize(n = n()) %>% pull()` unique sampling trips defined by unique combinations of 'site.id' and 'visit.no'.  There were `r dg %>% distinct(site.id) %>% summarize(n = n()) %>% pull()` waterbodies sampled, `r dg %>% filter(visit.no == 2) %>% summarize(n = n()) %>% pull()` of which were sampled twice. The primary response variables are the concentrations of CO~2~, CH~4~, and N~2~O dissolved in the surface water of each waterbody.  Concentration is expressed in both mass based units (umol/L, nmol/L: dissolved.co2, dissolved.ch4, dissolved.n2o) and as a saturation ratio (co2.sat.ratio, ch4.sat.ratio, n2o.sat.ratio).  Saturation ratio is defined as the ratio of the measured concentration to that expected if the waterbody were in equilibrium with the atmosphere.  A value of 1 indicates equilibrium, values < 1 indicate undersaturation and values > 1 indicate supersaturation.  Undersaturated waterbodies function as a sink for the atmospheric gas whereas supersaturated waterbodies function as a source.

@@

# Check CRS
st_crs(ecoR) # no EPSG code
ecoR <- st_transform(ecoR, 5070) # convert to CONUS Albers
st_crs(ecoR) # 5070
```

# Bubble plots

```{r}
# CO2 Bubble plots

ggplot() +
  geom_sf(data = ecoR, color = NA, aes(fill = WSA9_NAME)) +
  geom_sf(data = dg.sf, aes(size = dissolved.co2, color = co2.src.snk), 
          show.legend = "point") + # improve legend
  ggtitle("Dissolved CO2 (uM)") +
  scale_color_manual(values = c("white", "black"), name = "source/sink") +
  scale_size(name = "dissolved CO2 (uM)",
             range = c(0.1, 5), # custom size range
             breaks = c(10, 100, 300, 600)) # custom breaks
```

Spatial patterns in dissolved CH4 are not apparent, although all but two sites are supersaturated.

```{r}
ggplot() +
  geom_sf(data = ecoR, color = NA, aes(fill = WSA9_NAME)) +
  geom_sf(data = dg.sf, aes(size = dissolved.ch4, color = ch4.src.snk), 
          show.legend = "point") +
  ggtitle("Dissolved CH4 (uM)") +
  scale_color_manual(values = c("white", "black"), name = "source/sink") +
  scale_size(name = "dissolved CH4 (uM)",
             range = c(0.1, 10), # custom size range
             breaks = c(0.1, 1, 10, 50, 100)) # custom breaks
```

The prevalance of undersaturated dissolved N2O concentrations is an important finding.  The IPCC assumes surface waters are an important source of N2O to the atmosphere.   Webb et al. recently published a paper in PNAS reporting that 70 out of 101 reservoirs sampled in Saskatchewan were N2O sinks.  This paper made quite a splash.  In our study, `r sum(dg$n2o.src.snk == "sink")` out of `r sum(!is.na(dg$dissolved.n2o))` sampled waterbodies were an N2O sink. It is somewhat interesting to note that the some of the highest concentrations were observed in Indiana/Ohio, where I conducted much of my dissertation research on aquatic N2O emissions.

```{r}
ggplot() +
  geom_sf(data = ecoR, color = NA, aes(fill = WSA9_NAME)) +
  geom_sf(data = dg.sf, aes(size = dissolved.n2o, color = n2o.src.snk),
          show.legend = "point") +
  ggtitle("Dissolved N2O (nM)") +
  scale_color_manual(values = c("white", "black"), name = "source/sink") +
  scale_size(name = "dissolved N2O (nM)",
             range = c(0.1, 10), # custom size range
             breaks = c(1, 10, 25, 50, 100)) # custom breaks
```

We can also look at population estimates of N2O saturation ratio by ecoregion.  Wow, the mean N2O saturation is less than one for all ecoregions and the nation!  95% CI overlap one in only 3 levels of the stratification.  This is pretty big news. 
```{r}

# extract saturation indicator, order by n2o
pop.eco.sat <- pop.eco %>%
  filter(grepl("sat", indicator, ignore.case = TRUE)) %>%
  mutate(subpopulation = factor(subpopulation, 
                             levels = filter(pop.eco, indicator == "N2O_SAT_RATIO", subpopulation != "National") %>%
                               arrange(estimate) %>%
                               select(subpopulation) %>%
                               pull() %>%
                               c("National", .)))

# look at dissolved concentration indicator
ggplot(pop.eco.sat, aes(estimate, subpopulation)) +
  geom_point() +
  geom_errorbarh(aes(xmax = ucb95pct, xmin = lcb95pct)) +
  facet_wrap(~indicator, scales = "free_x") +
  theme(axis.title = element_blank())

```

@@

## Lagged scatterplots
Mike, can you integrate scatterplot code here?

## Semivariograms
Variograms display the variability among pairs of points at various separation distances.  I was unable to fit a model to the dissolved gas concentration variograms using the default fitting parameters in gstat.  The variograms suggest a lack of spatial structure, but I suspect the model parameterization can be improved.

```{r, warning=FALSE, message=FALSE}
# Variogram function requires sp object
dg.sp <- as(dg.sf, Class = "Spatial")

# CO2 variogram
co2.v <- variogram(dissolved.co2~1, dg.sp)
co2.v.fit <- fit.variogram(co2.v, vgm(c("Exp", "Mat", "Sph"))) # allow to choose best model.  Not sure which model warnings are associated with.
# co2.v.fit # chose sperical
co2.v.fit <- fit.variogram(co2.v, vgm("Sph")) # refit spherical, several warnings (?)

# CH4 variogram
ch4.v <- variogram(dissolved.ch4~1, dg.sp)
ch4.v.fit <- fit.variogram(ch4.v, vgm(c("Exp", "Mat", "Sph"))) # allow to choose best model.  Not sure which model warnings are associated with.
# ch4.v.fit # chose exp
ch4.v.fit <- fit.variogram(ch4.v, vgm("Exp")) # refit exponential, no convergence

# N2O variogram
n2o.v <- variogram(dissolved.n2o~1, dg.sp)
n2o.v.fit <- fit.variogram(n2o.v, vgm(c("Exp", "Mat", "Sph"))) # allow to choose best model.  Not sure which model warnings are associated with.
# n2o.v.fit # chose exp
n2o.v.fit <- fit.variogram(n2o.v, vgm("Exp")) # refit exponential, no convergence

# Plot variograms and models (if fit)
plot(co2.v, model = co2.v.fit, main = "CO2 semivariogram")
plot(ch4.v, main = "CH4 semivariogram")
plot(n2o.v, main = "N2O semivariogram")

```
